Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-09 paper/introduction.tex
Edits.

Consider the example shown in Fig.~\ref{fig:motivating-eg}.
This code represents the logic for a streaming query operator (also
known as a continuous query). The operator receives a continuous
stream of input messages, which are processed via a user-defined
function, and batched into time-windows associated with a logical
timestamp. Messages may be delivered out-of-order and multiple
messages with the same timestamp may be delivered. A timing-message,
periodically indicates that no more input messages with a timestamp
less than or equalt to a given value $t$ will be subsequently
delivered. At this point, the operator can create the output message
corresponding to time $t$ and deliver it to its successors.

>>>>

Consider the example shown in Fig.~\ref{fig:motivating-eg}.
This code represents the logic for a streaming query operator (also
known as a continuous query). The operator receives a continuous
stream of input messages, which are processed via a user-defined
function, and batched into time-windows associated with a logical
timestamp. Messages may be delivered out-of-order and multiple
messages with the same timestamp may be delivered. A timing-message,
periodically indicates that no more input messages with a timestamp
less than or equal to a given value $t$ will be subsequently
delivered. At this point, the operator can create the output message
corresponding to time $t$ and deliver it to its successors.

### 2015-11-09 paper/related.tex
Edits.

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically ensure the safety of
region-based memory management in ML. Following their seminal work,
static type systems for safe region-based memory management have been
extensively studied in the context of various languages and problem
settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
WW01, rust}. Our work differs from the existing proposals in a number
of ways. Firstly, our problem setting includes lexically scoped stack
regions and dynamic transferable regions (both programmer-managed) in
context of an object-oriented programming language equipped with
higher-order functions. Second, we adopt a two-pronged approach to
memory safety that relies on a combination of a simple static type
discipline and lightweight runtime checks. 
% The type system reduces the problem of ensuring safety of all memory
% accesses in a program into a problem of enforcing a state-transition
% discipline on memory accesses over a small subset of objects in the
% program. The later can be done efficiently at runtime. This lets us
% guarantee safety in presence of
In particular, our approach requires neither restrictive static
mechanisms (e.g., linear types and unique pointers) nor expensive
runtime mechanisms (e.g., garbage collection and reference counting)
in order to guarantee safety. Lastly, our region type system comes
equipped with full type inference that completely eliminates the need
to write region annotations on types to convince the type checker that
the program is safe.

@@

An ownership type system for safe region-based memory management in
real-time Java has been proposed by~\cite{MIT03}. Like us, they too
assume a source language with programmer-managed memory regions, and
focus on proving safety of programs written in that language. Their
source language admits various kinds of regions in order to support
shared-memory concurrency. However, all their regions have
lexically-scoped lifetimes. In contrast, we admit regions with
dynamically determined lifetimes in order to support message-passing
concurrency. We borrow outlives relation from their formal
development, and our type system bears some similarities to their's.
However, our language also admits parametric polymorphism (generics)
and higher-order functions, whose interaction is non-trivial in
context of a region type system. On the other hand, our type system
does not admit region kinds and ownership, and is considerably simpler
than theirs. Furthermore, our type system comes with a type
safety result that formalizes its guarantees with respect to a
well-defined operational semantics. Like Cyclone, ~\cite{MIT03}'s
language is explicitly typed. Although there is some support for local
type inference, region types for methods and classes need to be
written explicitly. In contrast, we support full type inference that
eliminates any such need.

~\cite{HMN01} proposes a flow-sensitive region-based memory management
for first-order programs that proposes to overcome some of the
drawbacks of~\cite{tofte97} by generalizing~\cite{tofte97}'s approach
to regions with dynamic lifetimes. However, dynamic regions are still
not first-class values of the language, and reference counting is
nonetheless needed to ensure memory safety.~\cite{WW01} extends lambda
calculus with first-class regions with dynamic lifetimes, and imposes
restrictive linear typing to control accesses to regions. All type
annotations need to be written explicitly. Moreover, invariants needed
to ensure safety under region transfers, such as the absence of
references that escape a transferable region, cannot be enforced
under~\cite{WW01}'s type system.

>>>>

Tofte and Talpin in~\cite{tofte93,tofte94,tofte97} introduce the
concept of a region type system to statically ensure the safety of
region-based memory management in ML. Following their seminal work,
static type systems for safe region-based memory management have been
extensively studied in the context of various languages and problem
settings~\cite{cyclone02, cyclone04, yates99, MIT03, DPJ09, HMN01,
WW01, rust, gpu14}. Our work differs from the existing proposals in a
number of ways. Firstly, our problem setting includes lexically scoped
stack regions and dynamic transferable regions (both
programmer-managed) in context of an object-oriented programming
language equipped with higher-order functions. Second, we adopt a
two-pronged approach to memory safety that relies on a combination of
a simple static type discipline and lightweight runtime checks. 
% The type system reduces the problem of ensuring safety of all memory
% accesses in a program into a problem of enforcing a state-transition
% discipline on memory accesses over a small subset of objects in the
% program. The later can be done efficiently at runtime. This lets us
% guarantee safety in presence of
In particular, our approach requires neither restrictive static
mechanisms (e.g., linear types and unique pointers) nor expensive
runtime mechanisms (e.g., garbage collection and reference counting)
in order to guarantee safety. Lastly, our region type system comes
equipped with full type inference that completely eliminates the need
to write region annotations on types to convince the type checker that
the program is safe.

@@

An ownership type system for safe region-based memory management in
real-time Java has been proposed by~\cite{MIT03}. Like us, they too
assume a source language with programmer-managed memory regions, and
focus on proving safety of programs written in that language. Their
source language admits various kinds of regions in order to support
shared-memory concurrency. However, all their regions have
lexically-scoped lifetimes. In contrast, we admit regions with
dynamically determined lifetimes in order to support message-passing
concurrency. We borrow outlives relation from their formal
development, and our type system bears some similarities to their's.
However, our language also admits parametric polymorphism (generics)
and higher-order functions, whose interaction is non-trivial in
context of a region type system. On the other hand, our type system
eschews the notions of region kinds and ownership, leading to a more
succinct formalization. Furthermore, we establish a type safety result
that formalizes its guarantees with respect to a well-defined
operational semantics. Like Cyclone, ~\cite{MIT03}'s language is
explicitly typed. Although there is some support for local type
inference, region types for methods and classes need to be written
explicitly. In contrast, we support full type inference that
eliminates any such need.

~\cite{HMN01} proposes a flow-sensitive region-based memory management
for first-order programs that proposes to overcome some of the
drawbacks of~\cite{tofte97} by generalizing~\cite{tofte97}'s approach
to regions with dynamic lifetimes. However, dynamic regions are still
not first-class values of the language, and reference counting is
nonetheless needed to ensure memory safety.~\cite{WW01} extends lambda
calculus with first-class regions with dynamic lifetimes, and imposes
linear typing to control accesses to regions. All type annotations
need to be written explicitly. Moreover, invariants needed to ensure
safety under region transfers, such as the absence of references that
escape a transferable region, cannot be enforced under~\cite{WW01}'s
type system.

The idea of using region-based memory management to facilitate the
safe transfer of rich data structures between computational nodes has
been previously explored by~\cite{gpu14} in the context of Scheme
language. However, their setting only includes lexically-scoped
regions for which Tofte and Talpin~\cite{tofte97} style analysis
suffices. In contrast, our language provides first-class support for
transferable regions with dynamic lifetimes. We require this
generality in order to support streaming query operators, such as the
one shown in Fig.~\ref{fig:motivating-eg}.
